Hamiltonian ABC

### 2015-06-08 uai-cr.tex
added refs

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{APPROXIMATE BAYESIAN COMPUTATION}\label{sec:abc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Consider the Bayesian inference task of either drawing samples from or learning an approximate model of the following (usually intractable) posterior distribution:
\begin{equation}
  \pi(\thetav | \y_1, \ldots, \y_N ) \propto \pi(\thetav) \pi( \y_1, \ldots, \y_N  | \thetav )
\end{equation}
where $\pi(\thetav)$ is a prior distribution over parameters $\thetav \in {\rm I\!R}^{D}$ and $\pi( \y_1, \ldots, \y_N  | \thetav )$ is the likelihood of $N$ data observations, where $\y_i \in {\rm I\!R}^J$.  In ABC, the vector of $J$ observations are typically informative statistics of the raw observations.  It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior.  
  The simulator is treated as generator of random pseudo-observations, i.e. $\x \simsim \pi( \x | \thetav )$ is a draw from the simulator.  Discrepancies between the simulator outputs $\x$ and the observations $\y$ are scaled by a closeness parameter $\eps$ and treated as likelihoods.  This is the equivalent to putting an $\eps$-kernel around the observations, and using a Monte Carlo estimate of the likelihood using $S$ draws of $\x$: 
\begin{equation}
  \pi_{\epsvec}( \y | \thetav ) =  \int \pi_{\epsvec}(\y | \x ) \pi( \x | \thetav ) d\x 
                           \approx  \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}(\y | \x^{(s)} ) \label{eq:abc_mc_approx}
\end{equation}

@@

Of particular concern to this paper is the behavior of the log-likelihoods for different values of $\epsilon$.  In the $\eps$-kernel case, the log-likelihood is very sensitive to small values of $\eps$:
\begin{eqnarray}
  \log \pi_{\epsvec}( \y | \thetav ) & = & \log \sum_s \mathcal{N}( \y | \x^{(s)}, \epsvec^2 ) \\
                          & = & \log \mathcal{N}( \y | \x^{(s)}, \epsvec^2 ) + \log\lp 1 + H \rp \\
                          & \approx & -\log \epsvec -\frac{1}{2 \epsvec^2} ( \y - \x^{(m)} )^2 
\end{eqnarray}
where $m$ is the simulation that is closest to $\y$, $H$ is a sum over terms close to $0$. We can see that the log-likelihood can be set arbitrarily small by decreasing $\eps$.  On the other hand, by using a model of the simulation at $\thetav$
\begin{eqnarray}
 \log \pi_{\epsvec}( \y | \thetav ) & \approx & -\frac{1}{2} \log (\sigma^2_{\thetav} + \epsvec^2 ) -\frac{( \y - \mu_{\thetav} )^2 }{2 (\sigma^2_{\thetav} + \epsvec^2 )} 
\end{eqnarray}
 For the SL model, $\epsvec$ acts as a smoothing term and can be set to small values with little change to the log-likelihood, as long as the SL estimators are fit appropriately.  This insensitivity to $\eps$ will be used in Section~\ref{sec:habc} for estimating gradients of the ABC likelihood.  Before describing HABC in full detail however, we now explain how scaling Hamiltonian dynamics in Bayesian learning can be accomplished using stochastic gradients from batched data.

@@

On the other hand, the synthetic-likelihood is stable; using a deterministic representation, 
\begin{equation}
\mu_{\thetav} = \frac{1}{S} \sum_s f( \thetav, \omega_s ) ~~~~~~~~~~~~~ \sigma^s_{\thetav} = \frac{1}{S-1} \sum_s ( \mu_{\thetav} - f( \thetav, \omega_s ) )^2
\end{equation}
the gradients (for a 1-dim problem) use $\epsvec$ as a smoothness prior in $\partial \lleps(\thetav)/\partial\thetav$:
\begin{equation}
  -\frac{1}{2}\log\lp\frac{\sigma_{\theta^+}^2 + \epsvec^2}{\sigma_{\theta^-}^2 + \epsvec^2}\rp -\frac{( \y - \mu_{\thetav^+} )^2 }{2 (\sigma^2_{\thetav^+} + \epsvec^2 )} +\frac{( \y - \mu_{\thetav^-} )^2 }{2 (\sigma^2_{\thetav^-} + \epsvec^2 )} 
\end{equation} 
In Figure~\ref{fig:exp-varg}, as part of our demonstration of HABC, we compare the gradient approximations around the true $\thetav_{\text{MAP}}$ using SL versus kernel-$\epsvec$ Figure~\ref{fig:exp-varg} for a simple problem.  We find that although, for this particular problem, SL has a small bias due to its Gaussian assumption, it has much smaller variance, an important property for HABC.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{DISCUSSION AND CONCLUSION} \label{sec:conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{-0.1in}
Hamiltonian ABC proposes a new set of algorithms for Bayesian inference of likelihood-free models.  HABC builds  upon the connections between Hamilton Monte Carlo with stochastic gradients and well-established gradient approximations based on a minimal number of forward simulations, even in high-dimensions.  We have performed some preliminary experiments showing the feasibility of running ABC on both small and large problems, and we hope that the door has been opened for exploration of larger simulation-based models using HABC.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{APPROXIMATE BAYESIAN COMPUTATION}\label{sec:abc}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Consider the Bayesian inference task of either drawing samples from or learning an approximate model of the following (usually intractable) posterior distribution:
\begin{equation}
  \pi(\thetav | \y_1, \ldots, \y_N ) \propto \pi(\thetav) \pi( \y_1, \ldots, \y_N  | \thetav )
\end{equation}
where $\pi(\thetav)$ is a prior distribution over parameters $\thetav \in {\rm I\!R}^{D}$ and $\pi( \y_1, \ldots, \y_N  | \thetav )$ is the likelihood of $N$ data observations, where $\y_i \in {\rm I\!R}^J$.  In ABC, the vector of $J$ observations are typically informative statistics of the raw observations.  It can be shown that if the statistics used in the likelihood function are sufficient, then these algorithms sample correctly from an approximation to the true posterior \cite{marin:2012}.  
  The simulator is treated as generator of random pseudo-observations, i.e. $\x \simsim \pi( \x | \thetav )$ is a draw from the simulator.  Discrepancies between the simulator outputs $\x$ and the observations $\y$ are scaled by a closeness parameter $\eps$ and treated as likelihoods.  This is the equivalent to putting an $\eps$-kernel around the observations, and using a Monte Carlo estimate of the likelihood using $S$ draws of $\x$: 
\begin{equation}
  \pi_{\epsvec}( \y | \thetav ) =  \int \pi_{\epsvec}(\y | \x ) \pi( \x | \thetav ) d\x 
                           \approx  \frac{1}{S} \sum_{s=1}^S \pi_{\epsvec}(\y | \x^{(s)} ) \label{eq:abc_mc_approx}
\end{equation}

@@

Of particular concern to this paper is the behavior of the log-likelihoods for different values of $\epsvec$.  In the $\epsvec$-kernel case, the log-likelihood is very sensitive to small values of $\epsvec$:
\begin{eqnarray}
  \log \pi_{\epsvec}( \y | \thetav ) & = & \log \sum_s \mathcal{N}( \y | \x^{(s)}, \epsvec^2 ) \\
                          & = & \log \mathcal{N}( \y | \x^{(s)}, \epsvec^2 ) + \log\lp 1 + H \rp \\
                          & \approx & -\log \epsvec -\frac{1}{2 \epsvec^2} ( \y - \x^{(m)} )^2 
\end{eqnarray}
where $m$ is the simulation that is closest to $\y$, $H$ is a sum over terms close to $0$. We can see that the log-likelihood can be set arbitrarily small by decreasing $\epsvec$.  On the other hand, by using a model of the simulation at $\thetav$
\begin{eqnarray}
 \log \pi_{\epsvec}( \y | \thetav ) & \approx & -\frac{1}{2} \log (\sigma^2_{\thetav} + \epsvec^2 ) -\frac{( \y - \mu_{\thetav} )^2 }{2 (\sigma^2_{\thetav} + \epsvec^2 )} 
\end{eqnarray}
 For the SL model, $\epsvec$ acts as a smoothing term and can be set to small values with little change to the log-likelihood, as long as the SL estimators are fit appropriately.  This insensitivity to $\epsvec$ will be used in Section~\ref{sec:habc} for estimating gradients of the ABC likelihood.  Before describing HABC in full detail however, we now explain how scaling Hamiltonian dynamics in Bayesian learning can be accomplished using stochastic gradients from batched data.

@@

On the other hand, the synthetic-likelihood is stable; using a deterministic representation, we have 
\begin{equation}
\mu_{\thetav} = \frac{1}{S} \sum_s f( \thetav, \omega_s ) ~~~~~~~~~~~~~ \sigma^s_{\thetav} = \frac{1}{S-1} \sum_s ( \mu_{\thetav} - f( \thetav, \omega_s ) )^2
\end{equation}
the gradients (for a 1-dim problem) use $\epsvec$ as a smoothness prior in $\partial \lleps(\thetav)/\partial\thetav$:
\begin{equation}
  -\frac{1}{2}\log\lp\frac{\sigma_{\theta^+}^2 + \epsvec^2}{\sigma_{\theta^-}^2 + \epsvec^2}\rp -\frac{( \y - \mu_{\thetav^+} )^2 }{2 (\sigma^2_{\thetav^+} + \epsvec^2 )} +\frac{( \y - \mu_{\thetav^-} )^2 }{2 (\sigma^2_{\thetav^-} + \epsvec^2 )} 
\end{equation} 
In Figure~\ref{fig:exp-varg}, as part of our demonstration of HABC, we compare the gradient approximations around the true $\thetav_{\text{MAP}}$ using SL versus kernel-$\epsvec$ Figure~\ref{fig:exp-varg} for a simple problem.  We find that although, for this particular problem, SL has a small bias due to its Gaussian assumption, it has much smaller variance, an important property for HABC.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{DISCUSSION AND CONCLUSION} \label{sec:conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{-0.1in}
Hamiltonian ABC proposes a new set of algorithms for Bayesian inference of likelihood-free models.  HABC builds  upon the connections between Hamiltonian Monte Carlo with stochastic gradients and well-established gradient approximations based on a minimal number of forward simulations, even in high-dimensions.  We have performed some preliminary experiments showing the feasibility of running ABC on both small and large problems, and we hope that the door has been opened for exploration of larger simulation-based models using HABC.
